---
title: "Excess mortality in Northern Ireland"
format: pptx
editor: visual
slide-level: 3
---

## All persons

```{r}
#| label: libraries
#| echo: false
#| message: false
#| warning: false

source(here::here("R/00_libraries.R"))

```


```{r}
#| echo: false
#| label: data

predictions <- read.csv(
  here::here("outputs/predictions/All Persons.csv")) |> 
  mutate(
    week_ending = as.Date(week_ending),
    across(c(areacode, age_group, sex, deprivation_quintile),
           \(x) factor(x))
  )

registered_deaths <- readxl::read_excel(
  here::here("data/Data for modelling - 2015-Qtr 3 2022.xlsx"),
  sheet = "All Persons"
) |> 
  rename(
    areacode = "Local Government District",
    deprivation_quintile = "MDM Quintile",
    week_ending = "Date",
    age_group = "Age_Group",
    registered = "Number of Deaths"
  ) |> 
  mutate(
    week_ending = as.Date(week_ending),
    across(c(areacode, age_group, sex, deprivation_quintile),
           \(x) factor(x))
  )

excess <- predictions |> 
  left_join(registered_deaths,
            by = c("areacode",
                   "age_group", 
                   "sex",
                   "deprivation_quintile",
                   "week_ending")) |> 
  rename(expected = "expected_deaths") |> 
  filter(
    week_ending <= as.Date("2022-09-30")
  ) |> 
  mutate(
    across(
      c(expected, registered),
      ~ case_when(
      is.na(.x) ~ 0,
      TRUE ~ .x
    )),
    sex = factor(sex,
                 levels = c("0", "1"),
                 labels = c("Male", "Female")),
    deprivation_quintile = factor(deprivation_quintile,
                                  levels = as.character(1:5),
                                  labels = c("Quintile 1 - most deprived",
                                             "Quintile 2",
                                             "Quintile 3",	
                                             "Quintile 4",
                                             "Quintile 5 - least deprived")),
    age_group = factor(age_group,
                       levels = c("85+", "75-84", "65-74", "50-64", "25-49", "0-24"))
  )

total_excess <- sum(excess$registered) - sum(excess$expected)

```

```{r}
#| label: chart-functions
#| echo: false

cumulative_excess <- function(data, y_axis, facet, title) {
  
  fill_colours <- c("neg" = "darkblue",
                    "pos" = "lightblue")

  cumulative_excess <- data |> 
    group_by(
      across(
        c({{ y_axis }}, {{ facet }})
      )
    ) |> 
    summarise(
      excess = sum(registered) - sum(expected),
      .groups = "drop") |> 
    mutate(
      sign = case_when(
        excess > 0 ~ "pos",
        TRUE ~ "neg"
        )) |> 
    ggplot(
      aes(
        x = excess,
        y = {{ y_axis }})
      ) +
    geom_col(
      aes(
        fill = sign),
      show.legend = FALSE) +
    scale_fill_manual(
      values = fill_colours
      ) +
    theme_bw() +
    labs(
      title = title,
      y = "",
      x = "Excess deaths"
      ) +
    facet_wrap(
      facets = vars({{ facet }})
    )
  
  return(cumulative_excess)
}


cumulative_ratio <- function(data, y_axis, facet, title, line_width = 20) {
  
  fill_colours <- c("neg" = "darkblue",
                    "pos" = "lightblue")

  cumulative_ratio_chart <- data |> 
    group_by(
      across(
        c({{ y_axis }}, {{ facet }})
      )
    ) |> 
    summarise(
      across(c(registered, expected),
             sum),
      .groups = "drop") |> 
    mutate(
      expected = case_when(
        expected < 1 ~ 0,
        TRUE ~ expected
      ),
      ratio = registered / expected,
      pos_neg = case_when(
        ratio > 1 ~ "pos",
        TRUE ~ "neg"
        )
      ) |>  
    ggplot(aes(y = {{ y_axis }},
               x = ratio)) +
    geom_segment(aes(
      yend = {{ y_axis }},
      x = 1,
      xend = ratio,
      colour = pos_neg),
      linewidth = line_width
      ) +
    scale_colour_manual(name = "",
                        values = fill_colours) +
    theme_bw() +
    theme(legend.position = "none") +
    labs(title = title,
         y = "", 
         x = "Ratio (registered / expected)") +
    scale_x_continuous(
      expand = expansion(mult = c(0.2, 0.1)
    )) +
    facet_wrap(
      facets = vars({{ facet }})
    )
  
  return(cumulative_ratio_chart)
}

```

For the period of week ending `r min(excess$week_ending)` to week ending `r max(excess$week_ending)` there were `r scales::comma(total_excess)` excess deaths.

---

### All persons weekly excess

```{r all-persons-weekly}
#| label: all-persons-weekly
#| echo: false

excess |> 
  group_by(week_ending) |> 
  summarise(
    across(
      c(registered, expected),
      \(x) sum(x, na.rm = TRUE)
    ),
    .groups = "drop"
  ) |> 
  mutate(excess = registered - expected,
         sign = case_when(
           excess > 0 ~ "pos",
           TRUE ~ "neg"
         )) |> 
  ggplot(
    aes(x = week_ending)
  ) +
  geom_col(
    aes(
      y = excess,
      fill = sign),
    show.legend = FALSE) +
  scale_fill_manual(
    values = c("pos" = "lightblue",
               "neg" = "darkblue")
  ) +
  theme_minimal() +
  labs(
    title = "Weekly all persons excess deaths in Northern Ireland",
    x = "Week ending",
    y = "Excess deaths"
  )

```

## Age and Sex

### Cumulative sex-age - Excess

```{r}
#| label: sex-age-excess


excess |> 
  cumulative_excess(
    y_axis = age_group,
    facet = sex,
    paste(
      "Cumulative excess deaths by age group and sex between week ending",
      paste(range(excess$week_ending),
            collapse = " and week ending ")
  ))
  

```

### Cumulative sex-age - Ratio of registered to expected

```{r}
#| label: sex-age-ratio


excess |> 
  cumulative_ratio(
    y_axis = age_group,
    facet = sex,
    paste(
      "Cumulative ratio of registered to expected deaths by\nage group and sex between week ending",
      paste(range(excess$week_ending),
            collapse = " and week ending ")
  ))
  

```

## Deprivation and Sex

### Cumulative sex-deprivation - Excess

```{r}
#| label: sex-deprivation-excess

excess |> 
  cumulative_excess(
    y_axis = deprivation_quintile,
    facet = sex,
    paste(
      "Cumulative excess deaths by deprivation quintile and sex\nbetween week ending",
      paste(range(excess$week_ending),
            collapse = " and week ending ")
  ))
  

```

### Cumulative sex-deprivation - Ratio of registered to expected

```{r}
#| label: sex-deprivation-ratio

excess |> 
  cumulative_ratio(
    y_axis = deprivation_quintile,
    facet = sex,
    title = paste(
      "Cumulative ratio of registered to expected deaths by\ndeprivation quintile and sex between week ending",
      paste(range(excess$week_ending),
            collapse = " and week ending ")),
    line_width = 24
  )
  

```

## Area and age

### Cumulative area-age - Excess

```{r}
#| label: area-age-excess

excess |> 
  cumulative_excess(
    y_axis = age_group,
    facet = areacode,
    paste(
      "Cumulative excess deaths by age group and geography between week ending",
      paste(range(excess$week_ending),
            collapse = " and week ending ")
  ))
  

```

### Cumulative area-age - Ratio of registered to expected

```{r}
#| label: area-age-ratio

excess |> 
  cumulative_ratio(
    y_axis = age_group,
    facet = areacode,
    title = paste(
      "Cumulative ratio of registered to expected deaths by\nage group and geography  between week ending",
      paste(range(excess$week_ending),
            collapse = " and week ending ")
      ),
    line_width = 5)
  

```


## Cause of death

```{r}
#| label: cause-data

files <- list.files(here::here(
  "outputs/predictions"
))

remove_files <- paste0(
  c(
    "All Persons",
    "Care Home",
    "Home",
    "Hospice",
    "Hospital",
    "Other"
    ),
  ".csv")

files <- files[!(files %in% remove_files)]
files <- setNames(
  here::here(
  paste0("outputs/predictions/",
         files)),
  nm = gsub(".csv", "", files))

cause_expected <- purrr::map_df(
  files,
  .f = read.csv,
  .id = "disease"
)

cause_registered <- setNames(
  names(files),
  nm = names(files)
) |> 
  purrr::map_df(
  ~ readxl::read_excel(
    here::here("data/Data for modelling - 2015-Qtr 3 2022.xlsx"),
    sheet = .x
  ),
  .id = "disease"
) |> 
  rename(
      areacode = "Local Government District",
      week_ending = "Date",
      age_group = "Age_Group",
      registered = "Number of Deaths"
    ) |> 
  mutate(week_ending = as.character(week_ending))


cause_excess <- cause_expected |> 
  left_join(cause_registered,
            by = c(
              "disease",
              "areacode",
              "age_group", 
              "sex",
              "week_ending")) |> 
  rename(
    expected = "expected_deaths") |> 
  filter(
    week_ending <= as.Date("2022-09-30")
  ) |> 
  mutate(
    across(
      c(expected, registered),
      ~ case_when(
      is.na(.x) ~ 0,
      TRUE ~ .x
    )),
    sex = factor(sex,
                 levels = c("0", "1"),
                 labels = c("Male", "Female")),
    age_group = factor(age_group,
                       levels = c("85+", "75-84", "65-74", "50-64", "25-49", "0-24"))
  )
```

### Cumulative cause of death and age - Excess

```{r}
#| label: cause-age-excess

cause_excess |> 
  cumulative_excess(
    y_axis = age_group,
    facet = disease,
    paste(
      "Cumulative excess deaths by age group and condition between week ending",
      paste(range(excess$week_ending),
            collapse = " and week ending ")
  ))
  

```

### Cumulative cause of death - Ratio of registered to expected

```{r}
#| label: cause-age-ratio

cause_excess |> 
  cumulative_ratio(
    y_axis = age_group,
    facet = disease,
    title = paste(
      "Cumulative ratio of registered to expected deaths by\nage group and condition between week ending",
      paste(range(excess$week_ending),
            collapse = " and week ending ")
      ),
    line_width = 3)
  

```


## Place of death

```{r}
#| label: place-data

files <- paste0(
  c(
    "Care Home",
    "Home",
    "Hospice",
    "Hospital",
    "Other"
    ),
  ".csv")

files <- setNames(
  here::here(
  paste0("outputs/predictions/",
         files)),
  nm = gsub(".csv", "", files))

place_expected <- purrr::map_df(
  files,
  .f = read.csv,
  .id = "place"
)

place_registered <- setNames(
  names(files),
  nm = names(files)
) |> 
  purrr::map_df(
  ~ readxl::read_excel(
    here::here("data/Data for modelling - 2015-Qtr 3 2022.xlsx"),
    sheet = .x
  ),
  .id = "place"
) |> 
  rename(
      areacode = "Local Government District",
      week_ending = "Date",
      age_group = "Age_Group",
      registered = "Number of Deaths"
    ) |> 
  mutate(week_ending = as.character(week_ending))


place_excess <- place_expected |> 
  left_join(place_registered,
            by = c(
              "place",
              "areacode",
              "age_group", 
              "sex",
              "week_ending")) |> 
  rename(
    expected = "expected_deaths") |> 
  filter(
    week_ending <= as.Date("2022-09-30")
  ) |> 
  mutate(
    across(
      c(expected, registered),
      ~ case_when(
      is.na(.x) ~ 0,
      TRUE ~ .x
    )),
    sex = factor(sex,
                 levels = c("0", "1"),
                 labels = c("Male", "Female")),
    age_group = factor(age_group,
                       levels = c("85+", "75-84", "65-74", "50-64", "25-49", "0-24"))
  )
```

### Cumulative place of death and age - Excess

```{r}
#| label: place-age-excess

place_excess |> 
  cumulative_excess(
    y_axis = age_group,
    facet = place,
    paste(
      "Cumulative excess deaths by age group and place of death between week ending",
      paste(range(place_excess$week_ending),
            collapse = " and week ending ")
  ))
  

```

### Cumulative place of death - Ratio of registered to expected

```{r}
#| label: place-age-ratio

place_excess |> 
  cumulative_ratio(
    y_axis = age_group,
    facet = place,
    title = paste(
      "Cumulative ratio of registered to expected deaths by\nage group and place of death between week ending",
      paste(range(place_excess$week_ending),
            collapse = " and week ending ")
      ),
    line_width = 8)
  

```
